---
title: "Measuring Biodiversity"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
bibliography: REFERENCES.bib
---

```{r}
#| label: DoNotModify
#| include: false
### Utilities for R. 
# Installation of R packages if necessary
install_packages <- function(packages) {
  invisible(
    sapply(
      packages, 
      FUN = function(package) {
        if (!package %in% installed.packages()[, 1]) {
          install.packages(
            package, 
            repos = "https://cran.rstudio.com/", 
            quiet = TRUE
          )
        }
      }
    )
  )
}

# Basic packages
install_packages(
  c(
    # R tools
    "knitr", "formatR", "learnr", "tidyverse", 
    # Scientific packages
    "divent", "ape", "cluster", "SPECIES", "fundiversity"
  )
)

# Chunk font size hook: allows size='small' 
# or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)

# Tidyverse default options
library("tidyverse")
# Set ggplotting to bw plot default, but with transparent background elements.  
theme_set(theme_bw(base_size = 12))
theme_update(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA)
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))
# learnr functions
library("learnr")
```

## Presentation

This is the practical for the Biodiversity Measurement course.

The [talk](https://1drv.ms/p/s!Amvhelr37CHjmolPddkD-D6YS7GGbw?e=cl8bLb)is online.

## Data

```{r}
#| label: setup
#| include: false
# Data available for all exercises

# Paracou6 : Paracou database. 
library("tidyverse")
paracou6 <- read_csv2("data/paracou6.csv")

library("divent")
# p6_abd: abundances object
p6_abd <- as_abundances(paracou6$spName, names = "Paracou 6")

# p6_phylo: phylogény built from the taxonomy
library("ape")
paracou6 %>%
  select(Family, Genus, spName) %>% 
  # Requires factors
  mutate_if(is.character, as.factor)  %>% 
  # Create the phylogeny
  {as.phylo(~Family/Genus/spName, data = ., collapse = FALSE)} %>% 
  # Branch lengths are identical
  compute.brlen(method = 1) %>% 
  # Turn multichotomies to dichotomies
  multi2di() %>%  
  # Merge branches without divergence 
  collapse.singles ->
  p6_phylo

# Z: Similarity matrix
traits <- read_csv2("data/traits.csv")
library("cluster")
# Row names are needed, ignore deprecation warning
rownames(traits) <- traits$spName
traits %>% 
  # Keep P6 species only
  filter(spName %in% paracou6$spName) %>% 
  # Keep trait values only
  select(-(Family:spName)) %>% 
  # Compute distance between rows
  daisy(metric = "gower") %>% 
  as.matrix() ->
  dist_species
# Dissimilarities
Z <- 1 - dist_species/max(dist_species)

# dfAbdP6SubPlots: abundances per subplot
paracou6 %>% 
  # Grouping
  group_by(SubPlot, spName) %>% 
  # Number of trees per species and subbplot
  summarize(abundance = n())  %>%  
  # A column per subplot
  pivot_wider(
    names_from = spName, 
    values_from = abundance, 
    values_fill  = 0
  ) %>% 
  # Rename the site column
  rename(site = SubPlot) %>% 
  # Name the subplots
  mutate(site = paste("Subplot", site)) %>% 
  as_abundances() ->
  p6_sub_abd
```

### Reading Paracou database

2016 inventory.

```{r}
#| label: Paracou6.csv
library("tidyverse")
paracou6 <- read_csv2("data/Paracou6.csv")
# Show the data
paracou6
```

### Map

The map of P6 wapas is made with **ggplot2**.

```{r}
#| label: P6_map
#| exercise: true
paracou6 %>%
  # Filter the wapa trees
  filter(Genus == "Eperua") %>%
  # Make a plot
  ggplot() + 
  # Points represent the trees
  geom_point(
    aes(x = Xfield, y = Yfield, size = CircCorr, color = Species)
  ) + 
  # Fixed coordinates for a map
  coord_fixed() + 
  # Color selection
  scale_color_brewer(palette = "Set1") +
  # Polish the text
  labs(
    x = "X", 
    y = "Y", 
    caption = "Paracou, plot 6", 
    size = "Circumference", 
    color = "Species"
  )
```

### Rank Abundance Curve

Make an object of class `abundances`.

```{r}
#| label: AbdP6
library("divent")
paracou6 %>%
  # A vector of characters is needed
  pull(spName) %>% 
  # Make it an abundances object
  as_abundances() ->
  p6_abd
```

Produce a Whittaker plot with ggplot, fitting a lognormal distribution.

```{r}
#| label: Whittaker
#| exercise: true
#| hint: true
p6_abd
```

```{r Whittaker-hint}
?species_distribution
```

## Species-Neutral Diversity

### Richness

Count the number of species from the abundance vector: remember to eliminate the zeros if there are any.

```{r}
#| label: Richness1
#| exercise: true
#| hint: true
p6_abd %>% 
  as.numeric() ->
  p6_abd_vector

```

```{r}
#| label: Richness1-hint
sum(p6_abd_vector > 0)
```

How many singletons and doubletons are there in the community?

```{r}
#| label: Singletons
#| exercise: true
#| hint: true

```

```{r}
#| label: Singletons-hint
sum(p6_abd_vector == 1)
```

Use the function `div_richness()` in *divent*.

```{r}
#| label: Richness2
#| exercise: true
# Works with both p6_abd and p6_abd_vector
p6_abd %>% 
  div_richness(estimator = "naive")
```

Estimate richness with the Chao and Jackknife estimators.

```{r}
#| label: Richness3
#| exercise: true
#| hint: true

```

```{r}
#| label: Richness3-hint
?div_richness
```

The internal algorithm for selecting the order of the Jackknife is that of @Burnham1978: the second-order Jackknife is chosen if the null hypothesis of equality of the first- and second-order estimators is rejected at the default risk threshold of 5% (which can be modified by the `jack_alpha` argument), which is the case here.
A higher order costs the estimator more variance: it is only justified if the estimated value is significantly different, which means that we can hope to reduce the bias (underestimation) due to an order that is too low.
The criteria of @Brose2003 suggest that the first-order Jackknife should be chosen because the proportion of singletons is less than one-third.
The two approaches almost always agree: this is one of the rare cases where the choice is ambiguous.

To force order 1, the argument `jack_max = 1` can be used.

Obtain the 95% confidence interval of the Jackknife 2 estimator with the package **SPECIES**.
The data must be abundance frequencies, i.e. the number of species represented by each number of individuals.

```{r}
#| label: SPECIES
#| exercise: true
library("SPECIES")
p6_abd_vector %>% 
  # Format for SPECIES
  abd_freq_count() %>% 
  as.matrix() %>% 
  # Estimate the jackknife and extract $CI from the list of results
  jackknife(k = 2) %>% 
  # Extract an item of the list (note the quotes)
  pluck("CI")
```

"lb" and "ub" are the upper and lower bounds.

```{r}
#| label: Richness-quiz
#| echo: false
quiz(
  question("What is the best estimator of the richness of plot 6?",
    answer("Chao"),
    answer("Jackknife 1"),
    answer("Jackknife 2"),
    answer("Jackknife 1 or 2, the choice is not clear",  correct = TRUE),
    answer("Jackknife 3"),
    allow_retry = TRUE,
    incorrect = "See [the  course](https://ericmarcon.github.io/MesuresBioDiv2/chap-MesuresNeutres.html#techniques-destimation-non-param%C3%A9trique) and try again.\nJackknife 1 is preferable according to Brose et al. (2003) but the calculation algorithm chooses Jackknife 2."
  ),
  question("Is the estimate robust?",
    answer("yes"),
    answer("no, if jackknife 2 is used", correct = TRUE),
    allow_retry = TRUE,
    incorrect = "See [the course](https://ericmarcon.github.io/MesuresBioDiv2/chap-MesuresNeutres.html#techniques-destimation-non-param%C3%A9trique) and try again."
  ),
  question("Why?",
    answer("The choice of the Jackknife order is unclear", correct = TRUE),
    answer("Richness is severely underestimated"),
    answer("The variance of the estimator increases with the order of the Jackknife", correct = TRUE),
    allow_retry = TRUE,
    incorrect = "See [the course](https://ericmarcon.github.io/MesuresBioDiv2/chap-MesuresNeutres.html#techniques-destimation-non-param%C3%A9trique) and try again."
  )
)
```

### Diversity

Estimate the asymptotic entropy and Shannon's diversity of plot 6 using the best estimator.
Use the `as_numeric` argument to obtain a number rather than a tibble.

```{r}
#| label: Shannon1
#| exercise: true
p6_abd %>% ent_shannon()
p6_abd %>% div_hill(q = 1)
```

Compare the diversity estimate with those of the ChaoJost estimator and the naive estimator.

```{r}
#| label: Shannon2
#| exercise: true
#| hint: true
p6_abd %>% div_hill(q = 1, estimator = "ChaoJost")
p6_abd %>% div_hill(q = 1, estimator = "naive")
```

The estimation bias decreases with the order of the diversity.
It is very low for Simpson's diversity: check this.

```{r}
#| label: Diversityq2
#| exercise: true
#| hint: true
p6_abd %>% div_hill(q = 2)
# Complete the code

```

The function `div_hill()` chooses the best default estimator, presently "UnveilJ".
The order of the Jackknife applied to the richness estimate can be forced with argument `jack_max`.

### Diversity profile

Plot the diversity profile of plot 6 using the best estimator.

```{r}
#| label: Profile1
#| exercise: true
p6_abd %>% 
  profile_hill() %>% 
  autoplot() +
  # Don't print the legend
  theme(legend.position = "none")
```

Compare with the naive profile.

```{r}
#| label: Profile2
#| exercise: true
p6_abd %>% 
  profile_hill(estimator = "naive") ->
  p6_profile_naive
p6_abd %>% 
  profile_hill() ->
  p6_profile_best
# Bind the profiles in a single tibble
p6_profile_naive %>% 
  bind_rows(p6_profile_best) %>% 
  # Plot it
  ggplot() +
  geom_line(
    aes(
      x = order,
      y = diversity,
      color = estimator
    )
  )
```

### Rarefaction and extrapolation

Diversity can be estimated at an arbitrary level of sampling effort or coverage rate.

Estimate the diversity observed with one hectare of inventory (the plot measures 6.25 ha).

```{r}
#| label: Leveln
#| exercise: true
p6_abd %>% 
  div_hill(q = 1, level = round(abd_sum(., as_numeric = TRUE) / 6.25))
# Estimate richness rather than Shannon

```

Estimate the sample coverage.

```{r Coverage}
#| label: Coverage
#| exercise: true
p6_abd %>% 
  coverage()
```

Estimate the diversity for a 99% sample coverage and compare it with the asymptotic diversity (sample coverage equal to 100%).
Estimate the richness to evaluate the number of species corresponding to the last percentage of coverage.

```{r}
#| label: LevelC
#| exercise: true
p6_abd %>% 
  div_hill(q = 1, level = .99)
p6_abd %>% 
  div_hill(q = 1)
# Estimate richness rather than Shannon

```

Plot the richness accumulation curve up to twice the size of the inventory.

```{r}
#| label: DivAC
#| exercise: true
#| exercise.timelimit: 120
p6_abd %>% 
  accum_hill(
    q = 1, 
    levels = seq(
      from = 1, 
      to = 2 * abd_sum(., as_numeric = TRUE), 
      by = 100
    )
  ) %>% 
  autoplot()
# Try q = 1 and q = 2, by = 500 to fasten computation
```

```{r}
#| label: Diversity-quiz
#| echo: false
quiz(
  question("Estimation bias decreases with the order of diversity",
    answer("True",  correct = TRUE),
    answer("False"),
    incorrect = "See [the course](https://ericmarcon.github.io/MesuresBioDiv2/entropie.html#estimation-de-lentropie)."
  ),
  question("Why?",
    answer("The impact of rare species on diversity increases with q"),
    answer("The impact of rare species on diversity decreases with q", correct = TRUE),
    incorrect = "See [the course](https://ericmarcon.github.io/MesuresBioDiv2/entropie.html#estimation-de-lentropie)."
  ),
  question("What is the estimation technique for the UnveilJ estimator?",
    answer("Probabilities are estimated using an elaborate model",  correct = TRUE),
    answer("Probabilities are estimated by multiplying frequencies by the coverage rate"),
    answer("The number of unobserved species is estimated by the Jackknife", correct = TRUE),
    answer("Their distribution is geometric but of little importance", correct = TRUE),
    allow_retry = TRUE,
    incorrect = "See [the course](https://ericmarcon.github.io/MesuresBioDiv2/entropie.html#estimation-de-lentropie)."
  )
)
```

## Phylogenetic Diversity

### Creation of a phylogeny

A phylogeny can be created from the taxonomy.

```{r}
#| label: p6Phylo
#| exercise: true
#| message: false
library("ape")
paracou6 %>%
  # Keep a single line per species
  group_by(Family, Genus, spName) %>% 
  # Count individuals (useless, but group_by needs summarise)
  summarise(abundance = n()) %>% 
  # Get rid of the groups
  ungroup() %>% 
  # Requires factors
  mutate_if(is.character, as.factor)  %>% 
  # Create the phylogeny
  {as.phylo(~Family/Genus/spName, data = ., collapse = FALSE)} %>% 
  # Branch lengths are identical
  compute.brlen(method = 1) %>% 
  # Turn multichotomies to dichotomies
  multi2di() %>%  
  # Merge branches without divergence 
  collapse.singles() ->
  p6_phylo
```

Plot the phylogeny.

```{r}
#| label: Phylogeny
#| exercise: true
#| hint: true
#| exercise.timelimit: 120

```

```{r phylogenie-hint}
#| label: Phylogeny-hint
p6_phylo %>% 
  plot(type = "fan", show.tip.label = FALSE) 
```

### Phylogenetic Diversity Profile

Estimate the Shannon phylogenetic diversity of plot 6 with the best estimator.

```{r , exercise=TRUE, exercise.timelimit=120}
#| label: phyShannon
#| exercise: true
#| exercise.timelimit: 120
p6_abd %>% 
  div_phylo(q = 1, tree = p6_phylo)
# Complete the code
```

Plot the phylogenetic diversity profile.

```{r}
#| label: phyProfil
#| exercise: true
#| exercise.timelimit: 120
p6_abd %>% 
  profile_phylo(tree = p6_phylo) %>% 
  autoplot()
# Choose the estimator

```

Compare PD and richness.

## References
