---
title: "Biodiversity Measurement course support"
format: html
editor: visual
---

```{r}
#| label: DoNotModify
#| include: false
### Utilities for R.

# Necessary packages
packages_required <- c(
  # R tools
  "knitr", "tidyverse", 
  # Scientific packages
  "divent", "vegan"
)
# Already installed?
packages_installed <- vapply(
  packages_required, 
  FUN = requireNamespace, 
  FUN.VALUE = TRUE, 
  quietly = TRUE
) 
# Install missing packages
install.packages(names(packages_installed)[!packages_installed])

# Chunk font size hook: allows size='small' 
# or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

## Motivation

This document contains code chunks to help you understanding the course.
Don't hesitate to modify them, add your own notes and code.

It comes with a practical you will run at home (`Biodiv-Practical.Rmd`).

## Species distributions

### Broken stick

Random distributions are created by `divent::rxxx()`functions.

```{r}
library("tidyverse")
library("divent")
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution
  distribution = "bstick"
) %>% 
  # Whittaker plot
  autoplot()
```

### Geometric

Random distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution
  distribution = "geom",
  prob_geom = 0.1
) %>% 
  # Whittaker plot
  autoplot()
```

### Lognormal

Random distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution
  distribution = "lnorm",
  sd_lnorm = 1
) %>% 
  # Whittaker plot
  autoplot()
```

Modify the standard deviation.

### Log-series

Random distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species is calculated
  # Distribution
  distribution = "lseries",
  fisher_alpha = 40
) %>% 
  # Whittaker plot
  autoplot()
```

### Estimate the parameters

Draw a very large lognormal distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees = max supported integer value
  size = .Machine$integer.max,
  # Number of species
  species_number = 300,
  # Distribution
  distribution = "lnorm",
  sd_lnorm = 1
) %>% 
  # Name it
  mutate(site = "Large log-normal") ->
  comm_large_lnorm
```

Sample it.

```{r}
# Draw 3000 trees from the large community
rcommunity(
  1,
  size = 3000,
  abd = as.numeric(comm_large_lnorm)
) %>% 
  # Name it
  mutate(site = "Small log-normal") ->
  comm_small_lnorm
# Make a single tibble with all data
comm_large_lnorm %>% 
  # Make a single tibble
  bind_rows(comm_small_lnorm) ->
  comm_lnorm
# Plot the distributions and fit a lognormal model
autoplot(comm_lnorm, fit_rac = TRUE, distribution = "lnorm")
# Estimate the parameters
fit_rac(comm_lnorm, distribution = "lnorm") 
```

## Species Accumulation Curves

### BCI data

Gather all subplots.

```{r}
library("vegan")
data(BCI)
BCI %>% 
  # Make an abundances object
  as_abundances() %>% 
  # Make all plot weights equals
  mutate(weight = 1) %>% 
  # Gather all 50 plots
  metacommunity() %>% 
  # Name it
  mutate(site = "BCI") -> 
  BCI_all
autoplot(BCI_all, fit_rac = TRUE, distribution = "lnorm")
fit_rac(BCI_all, distribution = "lnorm") 
```
Abundance frequency counts:
```{r}
BCI %>% 
colSums() %>% 
abd_freq_count() %>% 
arrange(desc(abundance))
```

### Accumulation curve

```{r}
# Number of trees in BCI
BCI_n <- abd_sum(BCI_all, as_numeric = TRUE)
# Asymptotic richness
BCI_richness <- div_richness(BCI_all, as_numeric = TRUE)
BCI_all %>% 
  accum_hill(
    levels = seq(from = 1, to = 2 * BCI_n, by = 10),
    # Run simulations to estimate uncertainty.
    # Start with 10 simulations, run 100 later
    n_simulations = 10,
  ) %>% 
  autoplot() +
  geom_vline(xintercept = BCI_n, linetype = 2) +
  geom_hline(yintercept = BCI_richness, linetype = 1)
```

## Richness

### Dummy communities

```{r}
tibble(
  sp_1 = c(10, 10, 10, 10),
  sp_2 = c(10, 11, 1, 0)
) %>% 
  as_abundances() %>% 
  print() ->
  comm_dummy
```

Richness:

sample:

```{r}
div_richness(comm_dummy, estimator = "naive")
```

- asymptotic:

```{r}
div_richness(comm_dummy)
```

### Paracou 6

Prepare the data: gather the four subplots.

```{r}
paracou_6_abd %>% 
  metacommunity() ->
  comm_p6
autoplot(comm_p6)
```

Abundance frequencies:

```{r}
comm_p6 %>% 
  as.numeric() %>% 
  abd_freq_count()
```

Estimators.

```{r}
div_richness(comm_p6, estimator = "naive") %>% 
  bind_rows(div_richness(comm_p6, estimator = "Chao1")) %>% 
  bind_rows(div_richness(comm_p6, estimator = "iChao")) %>% 
  # Default estimator
  bind_rows(div_richness(comm_p6, estimator = "jackknife")) %>% 
  bind_rows(div_richness(comm_p6, estimator = "jackknife", jack_max = 1)) %>% 
  bind_rows(div_richness(comm_p6, estimator = "rarefy" ))
```

### Interpolation / extrapolation

Diversity of 1 ha in Paracou:

```{r}
# Number of trees per hectare
p6_density <- abd_sum(comm_p6, as_numeric = TRUE) / 6.25
# Richness
comm_p6 %>% 
  div_richness(level = round(p6_density))
```
Diversity of 10 ha in Paracou:

```{r}
# Richness
comm_p6 %>% 
  div_richness(level = round(10 * p6_density))
```

## Shannon

### Information function

```{r}
ggplot() +
  geom_function(fun = \(p) {log(1 / p)})
```

Species information.

Information function.

```{r}
ggplot() +
  geom_function(fun = \(p) {p * log(1 / p)})
```

### Paracou 6

Estimators.

```{r}
ent_shannon(comm_p6, estimator = "naive") %>% 
  bind_rows(ent_shannon(comm_p6, estimator = "UnveilJ")) %>% 
  bind_rows(ent_shannon(comm_p6, estimator = "ChaoJost"))
```

Hill numbers.

```{r}
div_hill(comm_p6, q = 1, estimator = "naive") %>% 
  bind_rows(div_hill(comm_p6, q = 1, estimator = "UnveilJ")) %>% 
  bind_rows(div_hill(comm_p6, q = 1, estimator = "ChaoJost"))
```
