---
title: "Diversity and Distribution course support"
format:
    html:
      number-sections: true
      toc: true
editor: visual
# https://quarto.org/docs/computations/execution-options.html#output-options
execute:
  # show code chunk output
  include: true
  # Show the code in the output
  echo: true
  # Show messages
  message: true
  # Show warnings
  warning: true
  # Cache code results
  cache: true
bibliography: references.bib
---

```{r}
#| label: Install
#| include: false
### Utilities for R.

# Necessary packages
packages_required <- c(
  # R tools
  "knitr", "tidyverse", 
  # Scientific packages
  "ade4", "divent", "GoFKernel", "sads", "untb", "vegan"
)
# Already installed?
packages_installed <- vapply(
  packages_required, 
  FUN = requireNamespace, 
  FUN.VALUE = TRUE, 
  quietly = TRUE
) 
# Install missing packages
install.packages(names(packages_installed)[!packages_installed])

# Chunk font size hook: allows size='small' 
# or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

# Motivation

This document contains code chunks to help you understanding the course.
Don't hesitate to modify them, add your own notes and code.

It comes with a practical you will run at home (`Biodiv-Practical.Rmd`).

# Species distributions

## Figures

### Preston diagram

```{r}
library("tidyverse")
library("vegan")
data(BCI)
# Sum the 50 1-ha plots
BCI_50ha_abd <- colSums(BCI)
BCI_50ha_abd %>% 
  log10() %>% 
  hist(breaks = 15, main = "Preston diagram")
```

Probability densities

```{r}
BCI_50ha_abd %>% 
  tibble(abundance = .) %>% 
  ggplot() +
  # Bounded density
  geom_density(aes(x = abundance), bounds = c(0, Inf)) +
  scale_x_log10()
```

### Whittaker diagram

```{r}
library("divent")
BCI_50ha_abd %>% 
  # Declare it is an "abundances" object
  as_abundances() %>% 
  autoplot()
```

## Broken stick

Random distributions are created by `divent::rxxx()`functions.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution
  distribution = "bstick"
) %>% 
  # Whittaker plot
  autoplot()
```

## Geometric

Random distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution
  distribution = "geom",
  prob_geom = 0.1
) %>% 
  # Whittaker plot
  autoplot()
```

## Lognormal

Random distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution
  distribution = "lnorm",
  sd_lnorm = 1
) %>% 
  # Whittaker plot
  autoplot()
```

Modify the standard deviation.

## Log-series

Random distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species is calculated
  # Distribution
  distribution = "lseries",
  fisher_alpha = 40
) %>% 
  # Whittaker plot
  autoplot()
```

## Estimate the parameters

Draw a very large lognormal distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees = max supported integer value
  size = .Machine$integer.max,
  # Number of species
  species_number = 300,
  # Distribution
  distribution = "lnorm",
  sd_lnorm = 1
) %>% 
  # Name it
  mutate(site = "Large log-normal") ->
  comm_large_lnorm
```

Sample it.

```{r}
# Draw 3000 trees from the large community
rcommunity(
  1,
  size = 3000,
  abd = as.numeric(comm_large_lnorm)
) %>% 
  # Name it
  mutate(site = "Small log-normal") ->
  comm_small_lnorm
# Make a single tibble with all data
comm_large_lnorm %>% 
  # Make a single tibble
  bind_rows(comm_small_lnorm) ->
  comm_lnorm
# Plot the distributions and fit a lognormal model
autoplot(comm_lnorm, fit_rac = TRUE, distribution = "lnorm")
# Estimate the parameters
fit_rac(comm_lnorm, distribution = "lnorm") 
```

## Zero-sum multinomial distribution

BCI parameters $\theta$ and $m$ are estimated following @etienne2005 with the untb package.

```{r}
#| label: bci_params
#| echo: true
# BCI data
library("vegan")
data(BCI)
BCI_50ha <- colSums(BCI)
# Don't load untb because of conflicts with dplyr::count() and select()
# library("untb")
if (file.exists("data/BCI_logkda.RData")) {
  # Use the saved result if available to save
  # computation resources
  load("data/BCI_logkda.RData")
} else {
  BCI_50ha %>% 
  # name conflict with dplyr::count()
  untb::count() %>% 
  # requires pari/gp installed and in the search paths,
  # or use gp_binary argument
  untb::logkda() ->
  BCI_logkda
  # Save the result
  save(BCI_logkda, file = "data/BCI_logkda.RData")
}
BCI_50ha %>% 
  # name conflict with dplyr::count()
  untb::count() %>% 
  # Estimate theta and m
  untb::optimal.params(log.kda = BCI_logkda) %>% 
  print() ->
  BCI_params
```

The distribution of the local community is derived.
$J$ is that of BCI and $m$ was estimated previously.

```{r}
#| label: bci_sim_code
#| echo: true
J <- sum(BCI)
theta <- BCI_params[1]
m <- BCI_params[2]
library("sads")
# Number of species in the local community
Svolkov(theta, m, J) %>% 
  # Must be integers for the simulation
  round() %>% 
  # Draw the species abundances
  rvolkov(theta, m, J) %>% 
  # Make it an abundances object top plot it
  as_abundances() %>% 
  # Whittaker plot
  autoplot() ->
  BCI_sim_neutral
# BCI real distribution
BCI_50ha %>% 
  sort(decreasing = TRUE) %>% 
  as_tibble() %>% 
  mutate(rank = seq_len(n())) -> 
  BCI_tbl_50ha
# Plot altogether
BCI_sim_neutral +
  geom_line(
    data = BCI_tbl_50ha,
    aes(x = rank, y = value)
  )
```

# Species Accumulation Curves

## BCI data

Gather all subplots.

```{r}
library("vegan")
data(BCI)
BCI %>% 
  # Make an abundances object
  as_abundances() %>% 
  # Make all plot weights equals
  mutate(weight = 1) %>% 
  # Gather all 50 plots
  metacommunity() %>% 
  # Name it
  mutate(site = "BCI") -> 
  BCI_all
autoplot(BCI_all, fit_rac = TRUE, distribution = "lnorm")
fit_rac(BCI_all, distribution = "lnorm") 
```

Abundance frequency counts help read the figure.

```{r}
BCI %>% 
colSums() %>% 
abd_freq_count() %>% 
arrange(desc(abundance))
```

## Accumulation of subplots

Accumulate subplots in random order (Monte-Carlo).
Plot the expectation and the confidence interval of the number of species with respect to the number of samples.

```{r}
BCI %>% 
  specaccum() %>% 
  plot()
```

## Accumulation curve

Compute the expected accumulation curve of BCI up to twice the sampling effort, with its confidence interval.

```{r}
# Number of trees in BCI
BCI_n <- abd_sum(BCI_all, as_numeric = TRUE)
# Asymptotic richness
BCI_richness <- div_richness(BCI_all, as_numeric = TRUE)
BCI_all %>% 
  accum_hill(
    levels = seq(from = 1, to = 2 * BCI_n, by = 10),
    # Run simulations to estimate uncertainty.
    # Start with 10 simulations, run 100 later
    n_simulations = 10,
  ) %>% 
  autoplot() +
  geom_vline(xintercept = BCI_n, linetype = 2) +
  geom_hline(yintercept = BCI_richness, linetype = 1)
```

# Canonical lognormal

ggplot seems to bug with weighted, bounded density and log scale.

```{r}
# Most adundant species
BCI_50ha_mas <- max(BCI_50ha_abd)
BCI_50ha_abd %>% 
  tibble(abundance = .) %>% 
  ggplot() +
  # Bounded density
  geom_density(
    aes(x = abundance, color = "Species curve"), 
    bounds = c(0, Inf)
  ) +
  # Weighted density
  geom_density(
    aes(x = abundance, weight = abundance, color = "Individuals curve"), 
    bounds = c(min(BCI_50ha_abd), max(BCI_50ha_abd))
  ) +
  scale_x_log10()
```

Correct curves.

```{r}
library("GoFKernel")
plot(
  density.reflected(log10(BCI_50ha_abd), lower = 0),
  type = "l", 
  col = "red", 
  ylim = c(0, .8),
  main = "Canonical BCI",
  xlab = "log10(abundance)"
)
lines(
  density.reflected(
    log10(BCI_50ha_abd), 
    weights = BCI_50ha_abd, 
    lower = min(log10(BCI_50ha_abd)), 
    upper = max(log10(BCI_50ha_abd))
  )
)
abline(v = max(log10(BCI_50ha_abd)), lty = 2)

```

# Richness

## Dummy communities

Build dummy communities.

```{r}
tibble(
  sp_1 = c(10, 10, 1000, 1000),
  sp_2 = c(10, 11,    1,    0)
) %>% 
  as_abundances() %>% 
  print() ->
  comm_dummy
```

Sample richness.

```{r}
div_richness(comm_dummy, estimator = "naive")
```

Asymptotic richness.

```{r}
div_richness(comm_dummy)
```

## Paracou 6

Prepare the data: gather the four subplots.

```{r}
paracou_6_abd %>% 
  # Gather all sites
  metacommunity() %>% 
  # Rename "matacommunity"
  mutate(site = "Paracou 6") ->
  p6_abundances
autoplot(p6_abundances)
```

Abundance frequencies.

```{r}
p6_abundances %>% 
  as.numeric() %>% 
  abd_freq_count()
```

Estimators of richness.

```{r}
div_richness(p6_abundances, estimator = "naive") %>% 
  bind_rows(div_richness(p6_abundances, estimator = "Chao1")) %>% 
  bind_rows(div_richness(p6_abundances, estimator = "iChao")) %>% 
  # Default estimator
  bind_rows(div_richness(p6_abundances, estimator = "jackknife")) %>% 
  bind_rows(div_richness(p6_abundances, estimator = "jackknife", jack_max = 1))
```

## Interpolation / extrapolation

Diversity of 1 ha in Paracou.

```{r}
# Number of trees per hectare
p6_density <- abd_sum(p6_abundances, as_numeric = TRUE) / 6.25
# Richness
p6_abundances %>% 
  div_richness(level = round(p6_density))
```

Diversity of 10 ha in Paracou.

```{r}
# Richness
p6_abundances %>% 
  div_richness(level = round(10 * p6_density))
```

# Shannon

## Information function

Information brought by an individual: $I(p_s) = \ln(1 / p_s)$.

```{r}
ggplot() +
  # Increase the default number of points (101 to 1001) for precision
  geom_function(fun = \(p) {log(1 / p)}, n = 1001) +
  labs(x = "Probability", y = "Information")
```

Information brought by a species: $p_s \ln(1 / p_s)$.

```{r}
ggplot() +
  geom_function(fun = \(p) {p * log(1 / p)}, n = 1001) +
  labs(x = "Probability", y = "Information")
```

## Paracou 6

Estimators.

```{r}
ent_shannon(p6_abundances, estimator = "naive") %>% 
  bind_rows(ent_shannon(p6_abundances, estimator = "UnveilJ")) %>% 
  bind_rows(ent_shannon(p6_abundances, estimator = "ChaoJost")) %>% 
  print() ->
  p6_shannon
```

Effective number of species: $e^H$.

```{r}
p6_shannon %>% 
  mutate(diversity = exp(entropy))
```

## Dummy communities

Compare entropy.

```{r}
comm_dummy
ent_shannon(comm_dummy, estimator = "naive")
```

# Simpson

## Paracou 6

Estimators.

```{r}
ent_simpson(p6_abundances, estimator = "naive") %>% 
  bind_rows(ent_simpson(p6_abundances, estimator = "Lande")) %>% 
  bind_rows(ent_simpson(p6_abundances, estimator = "UnveilJ")) %>% 
  bind_rows(ent_simpson(p6_abundances, estimator = "ChaoJost")) %>% 
  print() ->
  p6_simpson
```

Effective number of species: $1 / (1 - E)$.

```{r}
p6_simpson %>% 
  mutate(diversity = 1 / (1 - entropy))
```

# Generalized entropy

## Information

Information brought by an individual: $I(p_s) = \ln_q(1 / p_s)$.

```{r}
ggplot() +
  # Increase the default number of points (101 to 1001) for precision
  geom_function(fun = \(p) {ln_q(1 / p, q = 0)}, n = 1001, aes(color = "0")) + 
  geom_function(fun = \(p) {ln_q(1 / p, q = 1)}, n = 1001, aes(color = "1")) + 
  geom_function(fun = \(p) {ln_q(1 / p, q = 2)}, n = 1001, aes(color = "2")) +
  scale_y_continuous(limits = c(0, 10)) +
  labs(x = "Probability", y = "Information", color = "q")
```

Information brought by a species: $p_s \ln_q(1 / p_s)$.

```{r}
ggplot() +
  # Increase the default number of points (101 to 1001) for precision
  geom_function(fun = \(p) {p * ln_q(1 / p, q = 0)}, n = 1001, aes(color = "0")) + 
  geom_function(fun = \(p) {p * ln_q(1 / p, q = 1)}, n = 1001, aes(color = "1")) + 
  geom_function(fun = \(p) {p * ln_q(1 / p, q = 2)}, n = 1001, aes(color = "2")) +
  labs(x = "Probability", y = "Information", color  = "q")
```

## Paracou 6

Diversity profile.

```{r}
p6_abundances %>% 
  profile_hill() %>% 
  autoplot() +
  # Don't print the legend
  theme(legend.position = "none")
```

Compare estimators.

```{r}
p6_abundances %>% 
  profile_hill(estimator = "UnveilJ") %>%
  # Use the site name to store the estimator's name
  mutate(site = "UnveilJ") %>% 
  bind_rows(
    p6_abundances %>% 
    profile_hill(estimator = "ChaoJost") %>% 
    mutate(site = "ChaoJost")
  ) %>% 
  bind_rows(
    p6_abundances %>% 
    profile_hill(estimator = "naive") %>% 
    mutate(site = "naive")
  ) %>% 
  autoplot() +
  labs(color  = "Estimator")
```

Extreme values.

The reciprocal of the Berger-Parker index [@Berger1970] is the rarity of the most abundant species $^{+\infty}D$.

```{r}
# q = 1000 is enough for infinity
div_hill(p6_abundances, q = 1000)
# Compare to 
p6_abundances %>% 
  # Extract proobabilities
  as_probabilities(as_numeric = TRUE) %>% 
  # Max value
  max() %>% 
  # Reciprocal
  `^`(-1)
  
```

## Dummy communities

Compare diversity

```{r}
comm_dummy
div_hill(comm_dummy, q = 1, estimator = "naive")
```

# Comparing diversity

## Asymptotic diversity

Compare asymptotic diversity of Paracou and BCI, extend $q$.

```{r}
# Orders from 0 to 4
q_seq_0_4 <- seq(from = 0, to = 4, by = 0.1)
profile_hill(p6_abundances, orders = q_seq_0_4) %>% 
  bind_rows(profile_hill(BCI_all, orders = q_seq_0_4)) %>% 
  autoplot()
```

## Standardized diversity

The sampling effort is very different: 50 ha in BCI, 6.25 ha in Paracou.
The asymptotic estimation is less reliable in Paracou.

Compare diversity at the same sampling effort, say 10 ha.

```{r}
# Number of trees in 10 ha
p6_n_10ha <- round(abd_sum(p6_abundances, as_numeric = TRUE) / 6.25 * 10)
BCI_n_10_ha <- round(sum(BCI_all) / 5)
```

Compare diversity of 10 ha.

```{r}
profile_hill(p6_abundances, level = p6_n_10ha) %>% 
  bind_rows(profile_hill(BCI_all, level = BCI_n_10_ha)) %>% 
  autoplot()
```

## Sample coverage

```{r}
coverage(p6_abundances) %>% 
  bind_rows(coverage(BCI_all))
```

Compare diversity at the same sample coverage [@Chao2012b], that of Paracou.

```{r}
reference_coverage <- coverage(p6_abundances, as_numeric = TRUE)
profile_hill(p6_abundances, level = reference_coverage) %>% 
  bind_rows(profile_hill(BCI_all, level = reference_coverage)) %>% 
  autoplot()
```

# Phylogenetic diversity

## Mock data

A simple community with 3 species.

```{r}
mock_3sp_abd %>% 
  as_abundances() %>% 
  mutate(site = "Mock 3 species") %>% 
  print() ->
  mock_3sp_abundances
```

```{r}
library("ape")
# Prepare the tree for convenience: make the branches twice longer
mock_3sp_tree$edge.length <- 2 * mock_3sp_tree$edge.length
plot(mock_3sp_tree, direction = "downwards", srt = 90)
axis(2)
```

PD is the total length of branches

```{r}
div_pd(mock_3sp_abd, tree = mock_3sp_tree)
```

Rao.

```{r}
ent_rao(mock_3sp_abundances, tree = mock_3sp_tree, estimator = "naive")
```

The tree is normalized to height = 1 by default.
Without normalization:

```{r}
ent_rao(
  mock_3sp_abundances, 
  tree = mock_3sp_tree, 
  estimator = "naive", 
  normalize = FALSE
)
```

## Paracou 6

`paracou_6_taxo` is a taxonomic tree: distance is 1 between species of the same genus, 2 between genera of the same family, 3 between families.

```{r}
plot(paracou_6_taxo, type = "fan", show.tip.label = FALSE) 
```

Calculate Rao's entropy and phylogenetic entropy of order 2

```{r}
ent_rao(p6_abundances, tree = paracou_6_taxo) %>% 
  bind_rows(
    ent_phylo(p6_abundances, tree = paracou_6_taxo, q = 2)
  )
```

Rao's entropy is not an effective number of species, but the average phylogenetic distance between two individuals.
The corresponding Hill number is:

```{r}
div_phylo(p6_abundances, tree = paracou_6_taxo, q = 2)
```

Diversity profile

```{r}
profile_phylo(p6_abundances, tree = paracou_6_taxo) %>% 
  autoplot()
```

# Functional diversity

## Mock data

The same mock data is used.
Distances between species is now assumed to be in the trait space.
Distances between species.

```{r}
mock_3sp_dist
```

Plot the trait space.

```{r}
library("ade4")
mock_pcoa <- dudi.pco(mock_3sp_dist, scannf = FALSE)
scatter(mock_pcoa, posieig = "none")
```

Similarity between species.

```{r}
# The dist object must be converted to a matrix
mock_3sp_Z <- 1 - as.matrix(mock_3sp_dist) / max(mock_3sp_dist)
mock_3sp_Z
```

Ordinariness.

```{r}
# Probabilities of species
mock_3sp_prob <- mock_3sp_abd / sum(mock_3sp_abd)
# Ordinariness of species
mock_3sp_O <- as.numeric(mock_3sp_Z %*% mock_3sp_prob)
# Compare probabilities and ordinariness
tibble(
  Probability = mock_3sp_prob,
  Rarity_p = 1 / mock_3sp_prob,
  Ordinariness = mock_3sp_O,
  Rarity_0 = 1 / mock_3sp_O
)
```

Diversity.

```{r}
profile_similarity(
  mock_3sp_abundances, 
  similarities = mock_3sp_Z, 
  estimator = "naive"
) %>% 
  autoplot()
```

Diversity of order 2 (i.e. Rao's).

```{r}
ent_similarity(
  mock_3sp_abundances, 
  similarities = mock_3sp_Z, 
  q = 2, 
  estimator = "naive"
)
```

## Paracou 6

The species traits are maximum height (95^th^ percentile) and seed mass, maximum height (95^th^ percentile) and seed mass.
A distance matrix was built (see home practical).
Distribution of distances between species.

```{r}
paracou_6_fundist %>% 
  as.numeric() %>% 
  tibble(Distance = .) %>% 
  ggplot() +
  geom_density(aes(x = Distance))
```

Diversity profile.

```{r}
p6_abundances %>% 
  profile_similarity(
    similarities = 1 - as.matrix(paracou_6_fundist) / max(paracou_6_fundist)
  ) %>% 
  autoplot()
```

## Intraspecific variability

Coordinates of species A, B and C in the trait space.

```{r}
mock_pcoa$li
```

Plot individuals in the trait space, with a small variability.

```{r}
# Create coordinates
tibble(
  Species = rep(names(mock_3sp_abd), times = mock_3sp_abd),
  x = rep(mock_pcoa$li[, 1], times = mock_3sp_abd),
  y = rep(mock_pcoa$li[, 2], times = mock_3sp_abd)
) ->
  mock_3sp_ind
# Choose variability
variability <- 0.1
# Map
mock_3sp_ind %>% 
  # Add variability
  mutate(
    x = x + rnorm(n(), mean = 0, sd = variability),
    y = y + rnorm(n(), mean = 0, sd = variability)
  ) %>% 
  ggplot(aes(x = x, y = y, color = Species)) +
  geom_point() +
  coord_fixed()
```

Diversity of the 6 individuals.
First, without variability.

```{r}
# Choose variability
variability <- 0
# Create the point set
mock_3sp_ind %>% 
  # Add variability
  mutate(
    x = x + rnorm(n(), mean = 0, sd = variability),
    y = y + rnorm(n(), mean = 0, sd = variability)
  )  %>% 
  # Distance matrix
  dist() %>% 
  as.matrix() ->
  mock_6ind_dist
# Delete the row and column name of the matrix (they are not species names)
rownames(mock_6ind_dist) <- colnames(mock_6ind_dist) <- NULL
# Total abundance
mock_6ind_n <- nrow(mock_6ind_dist)
# Similarity
mock_6ind_Z <- 1 - mock_6ind_dist / max(mock_6ind_dist)
# Abundances of individuals are 1, or any identical value
mock_6ind_abd <- rep(1, mock_6ind_n)
# Diversity
profile_similarity(
  mock_6ind_abd, 
  similarities = mock_6ind_Z,
  estimator = "naive"
) %>% 
  autoplot()
```

Change the variability (0.1 first) and run the code again.

# Diversity partitioning

## Mock data

Three habitats: A, B, C.

```{r}
matrix(
  c(
    0, 0, 0, 0, 0, 1, 1, 1, 1, 1,
    1, 1, 1, 1, 1, 1, 0, 1, 0, 0,
    0, 0, 0, 0, 0, 0, 0, 1, 1, 1
  ),
  nrow = 3, 
  byrow = TRUE
) %>% 
  as_abundances() %>% 
  mutate(
    site = paste("Habitat", LETTERS[1:3]),
    weight = 1
  ) ->
  mock_island_abd
```

Richness.

```{r}
div_richness(mock_island_abd, estimator = "naive")
div_richness(mock_island_abd, gamma = TRUE, estimator = "naive")
```

Multiplicative partitioning: 10 effective species ($\gamma$) = 5 effective species per habitat ($\alpha$) times 2 habitats ($\beta$).

```{r}
div_part(mock_island_abd, q = 0, estimator = "naive")
```

Additive partitioning: 10 effective species ($\gamma$) = 5 effective species per habitat ($\alpha$) + 5 more species ($\beta$).

## Paracou 6

The plot is made of 4 subplots.

```{r}
# Diversity profiles of the 4 subplots
paracou_6_abd %>% 
  profile_hill() %>% 
  autoplot()
```

### Partitioning

Function `div_part()` computes $\alpha$, $\beta$ and $\gamma$ diversity.

```{r}
div_part(paracou_6_abd, q = 1)
```

### Step by step

```{r}
# Alpha entropy
paracou_6_abd %>% 
  ent_tsallis(q = 1) %>% 
  # Mean entropy
  group_by(order) %>% 
  summarise(ent_alpha = mean(entropy)) %>% 
  # Diversity
  mutate(div_alpha = exp_q(ent_alpha, q = 1)) %>% 
  print() ->
  paracou_6_alpha
```

```{r}
# Gamma entropy
paracou_6_abd %>% 
  # Group the subplots
  metacommunity() %>% 
  ent_tsallis(q = 1) %>%
  # Rename entropy for clarity
  rename(ent_gamma = entropy) %>% 
  # Diversity
  mutate(div_gamma = exp_q(ent_gamma, q = 1)) %>% 
  print() ->
  paracou_6_gamma
```

```{r}
# Make a single tibble with alpha and gamma diversities
paracou_6_alpha %>% 
  bind_cols(
    paracou_6_gamma %>% 
      select(ent_gamma, div_gamma)
  ) %>% 
  mutate(
    # beta entropy
    ent_beta = ent_gamma - ent_alpha,
    # beta diversity
    div_beta = div_gamma / div_alpha
  )
```

### Profiles

Not implemented yet in divent.

alpha diversity.

```{r}
# Values of q used for computation: default values of profile_xxx()
orders <- seq(from = 0, to = 2, by = 0.1)
sapply(
  # Apply ent_tsallis to all orders
  orders,
  FUN = \(q) {
    # Get a number to obtain a matrix with sapply
    ent_tsallis(paracou_6_abd, q = q, as_numeric = TRUE)
  }
) %>% 
  # Each row contains a subbplot, each column an order, values are entropy
  # Calculate mean entropy (weights are equal)
  colMeans() %>% 
  # Diversity
  exp_q(q = orders) %>% 
  # Make a tibble
  tibble(
    order = orders,
    div_alpha = .
  ) %>% 
  print() ->
  paracou_6_alpha_profile
```

gamma diversity.

```{r}
paracou_6_abd %>% 
  profile_hill(gamma = TRUE) %>% 
  # Rename diversity to recongnize gamma later
  rename(div_gamma = diversity) %>% 
  print() ->
  paracou_6_gamma_profile
```

Compute beta.

```{r}
paracou_6_gamma_profile %>% 
  # Join the tibbles by identical order
  inner_join(paracou_6_alpha_profile) %>% 
  # Calculate beta diversity
  mutate(div_beta = div_gamma / div_alpha) %>% 
  print() ->
  paracou_6_profile
```

Plot.

```{r}
paracou_6_profile %>% 
  # paracou_6_profile is an object of class profile that can be plotted
  # The column to plot must be named diversity
  rename(diversity = div_alpha) %>% 
  autoplot() + 
  theme(legend.position = "none")
```

```{r}
paracou_6_profile %>% 
  # rename div_beta to plot it
  rename(diversity = div_beta) %>% 
  autoplot() + 
  theme(legend.position = "none")
```
